---
title: "Term paper for crosslinguistic corpus linguistics"
author: "Claudia Beatriz Castillo Upiachihua, Student ID 7416549"
date: "2024-02-11"
output: 
  html_document:
    toc: true
bibliography: bibliography_citedrive.bib
---

# A corpus of Peruvian Spanish with a focus on word order

## I. Introduction

Every language unique and serves as a window into the human cognition, culture, and society. A single language could contain features that other language may not. Yet, all languages serve the universal purpose of communication. The study of a language not only unravels the structure and patterns embedded within linguistic expressions but also sheds light on the diverse ways in which language is used and understood across different contexts and communities. Thus, if one wants to study a language, large data of the language is needed. 

Spanish is a language spoken by millions all over the world. It exhibits considerable diversity across different regions, influenced by historical, social, geographical, and linguistic contact. Each Spanish-speaking country, and even within countries, regions, or communities, may have distinct linguistic features, known as dialects. The linguistic variation of these dialects manifests in differences in vocabulary, pronunciation, grammar, discourse patterns, etc. Among the Spanish-speaking countries, Peru stands out as a compelling case study for investigating linguistic diversity and variation. Situated in the heart of South America, Peru boasts a rich tapestry of cultural and linguistic heritage, shaped by indigenous languages, colonial history, and other sociopolitical factors. Within Peru itself, linguistic diversity is further amplified by the presence of multiple indigenous languages alongside Spanish, as well as regional variations in Spanish dialects. The Andean Spanish of Peru is a great example of linguistic diversity and how a language can be shaped when two to more languages interact. The Andean Spanish has been greatly influenced by the indigenous languages of the region not only phonologically speaking but also grammatically. It is known that this dialect makes use of OV structures more frequently than the coastal dialect, which is a typical feature of Quechua and Aymara languages that are spoken in this region[@Klee2011].

In linguistics, the creation and analysis of corpora offer invaluable insights into the intricate nuances of language structure, variation, and usage. Therefore, I decided to create a corpus of the region of the coast and analyse the frequency of OV sentences from the coast and see to what extend the OV sentences typical of andean Spanish are present in the coastal peruvian Spanish. My idea in the beginning was to collect data from Tiktok from 3 different regions of Peru, namely, data from the coastal, andean and jungle dialect to then compare their different word order. Nonetheless, there was not enough videos available online for this kind of research. Then I found a large dataset online and decided to work with this data annotate it and analyse it. The main aim of this paper is to build a corpus using a dataset [@guevara-rukoz-etal-2020-crowdsourcing] containing only sentences in the Peruvian standard dialect of Spanish. This data has been taken from different audios. It has been transcribed and corrected. The sentences will be annotated so that it will be easier to investigate different linguistic phenomena. Nevertheless, we will focus on the word order of Spanish. In addition, this paper aims to investigate the frequency of OV sentences on our dataset.

### Research question

## II. Literature Review 

The corpus approach has gain popularity in recent years as it is powerful methodology for analyzing different linguistic phenomena and linguistic variation. In this section, the importance of corpus-based studies to investigate linguistic phenomena will be highlighted, and a review of the literature on Peruvian Spanish word order will be provided as well.

Creating a corpus does not only involves collecting large bodies of text or speech, but also the annotation and analysis of this data, which helps to gain insights into language structure, usage, and variation [@biber1998]. Corpora serve as great and large resources for linguists, providing authentic language data that can be systematically examined to identify linguistic patterns and trends [@mcenery2019].

Regional corpora on the other hand play a crucial role in capturing the linguistic diversity and variation inherent within specific geographical contexts. By focusing on a particular region or dialect, unique linguistic features and language change can be revealed[@szmrecsanyi2017]. In the context of Peru, the coastal region exhibits a distinctive linguistic variety influenced by historical, social, and geographical factors. I will examine the distribution of word order in Peruvian coastal Spanish, But first it will be necessary to review existing literature on word order.

Spanish is a language that allows different positions of its constituencies. Although its canonical word order is SVO, it is more flexible than others SVO languages since its constituencies can have different positions within a sentence. The constituencies in a transitive sentences with a verb such as "Comprar" can take different places as the examples below taken from[@olarrea2012] paper. It might even seem like word order is rather free. However, some constituencies differ in order due to the discourse informational content, pragmatic functions, prosodic properties, such as new or previously mentioned information, and focus it wants to convey. And not all different word orders from the examples can be applied interchangeably within a specific discourse context. In other words, a sigle position of the elements provides information on its discourse context and or pragmatic function.

Examples from @olarrea2012

a\. (Él) compraba el periódico todos los días (SV0)

(He) used-to-buy the newspaper every day 

b\. (Él) compraba todos los días el periódico(SVO)

c\. El perio´dico, (el) lo compraba todos los dias (OSV)

The newspaper, (he) cl-it used-to-buy everyday

d\. Él, el periódico lo compraba todos los días (SOV)

e\. El periódico lo compraba (él) todos los días (OVS)(OV)

f\. El periódico es lo que (él) compraba todos los días (OSV)(OV)

The newspaper is what (he) used-to-buy everyday

g\. El periódico es lo que compraba todos los días (él) (OVS) (OV)

h\. Compraba (él) el periódico todos los días ()

i\. Todos los dı´as compraba (él) el perio´dico.

According to @ocampo1995, In standard Spanish the object-verb word order is pragmatically marked, and it is used to indicate a contrary to expectation function, contrast, focal constituent and topic. This word order occurs no so frequent in non-contact Spanish such as "limeño" Spanish. Another study by @KleeandOcampo that investigated the pragmatic functions of OV and VO constructions in Calca Spanish discovered that the speakers make use of the object-verb word order to express repetition, summary, agreement and explanation in addition to the previously mentioned pragmatic functions.

SVO is the canonical word order in Spanish. However When the subject is narrowly focused, it typically appears at the end of the sentence creating a VOS order. We can then say that focus marking affects the sentence structure, which often results in the alteration of the canonical SVO order(Domínguez2013)

One crucial property of Spanish is that the ordering of elements in a sentence reveals a particular information structure which is unique to that utterance. We can say that focus marking affects the structure of sentences, which often results in the alteration of the canonical SVO order

The F-marked subject (Marta) is the only element bearing main stress in sentence-final position in, a position that is unusual for subjects in languages with canonical SVO order (Zubizarreta. This example shows that there is a narrow link between where stress in assigned and where narrowly-focused elements can appear in a sentence in Spanish.

## III. Methodology

```{r libraries}
library(tidyverse)
library(dplyr)
```

Questions: 
How to know the word order of the sentences from dependancy tree?



```{r datasets}

documents <- read.csv("../transcriber/output/parsed/documents.csv")
sentences <- read.csv("../transcriber/output/parsed/sentences.csv")
tokens <- read.csv("../transcriber/output/parsed/tokens.csv")

```

```{r get word order function}

get_sentences_word_order <- function(sentences_in) {
  
  word_orders <- tibble(
    doc_id=character(),
    doc_title=character(),
    sentence_id=character(),
    sentence_index=character(),
    verb_id=character(),
    verb_index=character(),
    word_order=character(), 
    dialect=character())
  
    for (i in 1:nrow(sentences_in)) {
      sentence <- sentences_in[i, ]
      
      document <- (documents %>% filter(id == sentence$doc_id))[1, ]
    
      sentence_tokens <- tokens %>% filter(sentence_id == sentence$id)
    
      sentence_verb_tokens <- sentence_tokens %>% filter(upos == 'VERB')
      
      if (nrow(sentence_verb_tokens) > 0) {
        for (i in 1:nrow(sentence_verb_tokens)) {
          verb <- sentence_verb_tokens[i, ]
          nsubj <- sentence_tokens %>% filter(head == verb$index & deprel == 'nsubj')
          obj <- sentence_tokens %>% filter(head == verb$index & deprel == 'obj')
    
          if (nrow(nsubj) == 1 & nrow(obj) == 1) {
    
          # create named vector with s/v/o as names and their indices as value
            word_order_pos <- c('s' = nsubj[1, ]$index, 'v' = verb$index, 'o' = obj[1, ]$index)
          # sort by indices then get names
            sorted_word_order <- names(sort(word_order_pos))
          # concatenate
            word_order <- paste(sorted_word_order, collapse = '')
    
            word_orders <- word_orders %>%  add_row(
              doc_id=document$id,
              doc_title=document$title,
              sentence_id=sentence$id,
              sentence_index=as.character(sentence$index),
              verb_id=verb$id,
              verb_index=verb$index,
              word_order=word_order,
              dialect=document$dialect)
          }
        }
      }
    }
  
    return(word_orders)

}
```

## IV. Data analysis
```{r}
left_join(sentences, documents %>% rename(doc_id = id), by = "doc_id") %>%
  group_by(dialect) %>% 
  summarise(count = n())

```

```{r word order for all sentences by dialect}
get_sentences_word_order(sentences) %>%
  group_by(dialect, word_order) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = word_order, y = count, fill = dialect)) + 
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(x = "Word Order", y = "Count",
       title = "Word Orders for All Sentences by Dialects")
```

```{r word order for affirmative sentences by dialect}
affirmative_sentence_ids <- (tokens %>% group_by(sentence_id) %>% filter(!any(text == '?')) %>% summarise())$sentence_id
affirmative_sentences <- sentences %>% filter(id %in% affirmative_sentence_ids)

get_sentences_word_order(affirmative_sentences) %>%
  group_by(dialect, word_order) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = word_order, y = count, fill = dialect)) + 
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(x = "Word Order", y = "Count",
       title = "Word Orders for Affirmative Sentences by Dialects")

```

## V. Discussion

## VI. Conclusion

## References
